---
title: "haplo_validate"
author: "Changyi"
date: "2026-06-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

haplotype reconstruction
This is the script to reconsruct the haplotype data


```{r}
library(haploReconstruct)
library(haplovalidate)
library(data.table)
library(poolSeq)
library(ggplot2)
library(ggpubr)
```

#1. data preparation
only significant SNPs are required for haplovalidate to cluster
1) AF for the significant SNPs
2) CMH results
```{r}
num_repl <- 5
gen <- c(10,20,30,40,50,60)
len_2L=as.integer(21106324)
len_3L=as.integer(22263419)

sync_colname<-c("chr","pos","REF",
           paste0("anc_F0_r",1:5),
           paste0("cons_F",rep(gen, each = num_repl),"_r",rep(1:num_repl, length(gen))),
           paste0("cons_F160_r",1:num_repl),
           paste0("fluc_F",rep(gen, each = num_repl),"_r",rep(1:num_repl, length(gen))),
           paste0("fluc_F160_r",1:num_repl)
           )

fluc_colname<-c("chr","pos","REF",
           paste0("anc_F0_r",1:5),
           paste0("fluc_F",rep(gen, each = num_repl),"_r",rep(1:num_repl, length(gen)))
           )

cons_colname<-c("chr","pos","REF",
           paste0("anc_F0_r",1:5),
           paste0("cons_F",rep(gen, each = num_repl),"_r",rep(1:num_repl, length(gen)))
           )


dt_sync<-fread("../../00data/genomic_data/F0_anc.F10-F60_F160_cons.F10-F60_F160_fluc.5rep.sync.gz",col.names = sync_colname)


#fluc
cmh_fluc<-fread("../../00data/genomic_data/fluctuating_F0_F60_CMH.result.gz")
cmh_fluc<-cmh_fluc[adj_p<0.05,]
dt_sync_fluc_sig<-dt_sync[cmh_fluc, on = .(chr, pos), nomatch = 0][, ..fluc_colname]
fwrite(dt_sync_fluc_sig, file = "Fluc_F0_F60_cmh_sig_snps.sync", quote = FALSE, sep="\t", col.names = FALSE)

cmh_fluc[, score := -log10(adj_p)]
cmh_fluc<-cmh_fluc[chr!="4",]
invisible(cmh_fluc[, new_chr := fcase(
  chr %chin% c("2L","2R"), "2",
  chr %chin% c("3L","3R"), "3",
  default = "X"
)])
invisible(cmh_fluc[, new_pos := fcase(
  chr=="2R",pos + len_2L,
  chr=="3R", pos + len_3L,
  default = pos
)])
cols_to_keep <- c("new_chr","new_pos","score")
cmh_fluc<-cmh_fluc[,..cols_to_keep]
setnames(cmh_fluc, c("chr","pos","score"))
saveRDS(cmh_fluc, file = "Fluc_cmh_p.rds")
#cons
cmh_cons<-fread("../../00data/genomic_data/constant_F0_F60_CMH.result.gz")
cmh_cons<-cmh_cons[adj_p<0.05,]
dt_sync_cons_sig<-dt_sync[cmh_cons, on = .(chr, pos), nomatch = 0][, ..cons_colname]
fwrite(dt_sync_cons_sig, file = "Cons_F0_F60_cmh_sig_snps.sync", quote = FALSE, sep="\t", col.names = FALSE)

cmh_cons[, score := -log10(adj_p)]
cmh_cons<-cmh_cons[chr!="4",]
invisible(cmh_cons[, new_chr := fcase(
  chr %chin% c("2L","2R"), "2",
  chr %chin% c("3L","3R"), "3",
  default = "X"
)])
invisible(cmh_cons[, new_pos := fcase(
  chr=="2R",pos + len_2L,
  chr=="3R", pos + len_3L,
  default = pos
)])
cols_to_keep <- c("new_chr","new_pos","score")
cmh_cons<-cmh_cons[,..cols_to_keep]
setnames(cmh_cons, c("chr","pos","score"))
saveRDS(cmh_cons, file = "Cons_cmh_p.rds")
```

##1.1 extract AF information
```{r}
repl <- 1:5
gens <- c(0,10,20,30,40,50,60)

len_2L=as.integer(21106324)
len_3L=as.integer(22263419)


#fluc
sync_fluc<-read.sync("Fluc_F0_F60_cmh_sig_snps.sync", 
                     gen = rep(gens,each=length(repl)), 
                     repl = rep(repl,length(gens)), polarization = "rising")
info_fluc<-sync_fluc@alleles
#exclude chr4
info_fluc<-info_fluc[chr!="4",]
invisible(info_fluc[, fallallele := fifelse(rising != ref, ref,
                                  fifelse(major == ref, minor, major))])
invisible(info_fluc[, basePops := rowMeans(.SD, na.rm = TRUE), .SDcols = paste0("F0.R",1:5,".freq")])
invisible(info_fluc[, new_chr := fcase(
  chr %chin% c("2L","2R"), "2",
  chr %chin% c("3L","3R"), "3",
  default = "X"
)])
invisible(info_fluc[, new_pos := fcase(
  chr=="2R",pos + len_2L,
  chr=="3R", pos + len_3L,
  default = pos
)])

cands_col<-c("new_chr","new_pos","ref","rising","fallallele", "basePops",
             paste0("F",rep(gens, each=length(repl)),".R",rep(repl,length(gens)),".freq"))
cands_fluc <- info_fluc[, ..cands_col]
cands_fluc[, new_chr := factor(new_chr, levels = c("2","3","X"))]
setorder(cands_fluc, new_chr, new_pos)
setnames(cands_fluc, c("chr","pos","ref","riseallele","fallallele","basePops",
                       paste0("L",1:35)))
saveRDS(cands_fluc, file = "Fluc_F0_F60_cmh_sig_AF.rds")

#cons
sync_cons<-read.sync("Cons_F0_F60_cmh_sig_snps.sync", 
                     gen = rep(gens,each=length(repl)), 
                     repl = rep(repl,length(gens)), polarization = "rising")
af_cons<-af(sync_cons, gen = gens, repl = repl)
info_cons<-sync_cons@alleles
#exclude chr4
info_cons<-info_cons[chr!="4",]
invisible(info_cons[, fallallele := fifelse(rising != ref, ref,
                                  fifelse(major == ref, minor, major))])
invisible(info_cons[, basePops := rowMeans(.SD, na.rm = TRUE), .SDcols = paste0("F0.R",1:5,".freq")])
invisible(info_cons[, new_chr := fcase(
  chr %chin% c("2L","2R"), "2",
  chr %chin% c("3L","3R"), "3",
  default = "X"
)])
invisible(info_cons[, new_pos := fcase(
  chr=="2R",pos + len_2L,
  chr=="3R", pos + len_3L,
  default = pos
)])
cands_col<-c("new_chr","new_pos","ref","rising","fallallele", "basePops",
             paste0("F",rep(gens, each=length(repl)),".R",rep(repl,length(gens)),".freq"))
cands_cons <- info_cons[, ..cands_col]

setnames(cands_cons, c("chr","pos","ref","riseallele","fallallele","basePops",
                       paste0("L",1:35)))
cands_cons[, chr := factor(chr, levels = c("2","3","X"))]
setorder(cands_cons, chr, pos)
saveRDS(cands_cons, file = "Cons_F0_F60_cmh_sig_AF.rds")

```





#2.perform haplotype clustering
### 2.1 initialize parameters
```{r}
 repl <- 1:5
 gens <- c(0,10,20,30,40,50,60)

 base.pops <- c(rep(TRUE, length(repl)),rep(FALSE,length(repl)*(length(gens)-1)))
 polaRise <- list(c(1,31),c(2,32),c(3,33),c(4,34),c(5,35)) 
 compare <- rep(TRUE,length(repl)*length(gens))
 pop.ident <- rep(repl,length(gens))
 pop.generation  <- rep(gens, each=length(repl))

```

###2.2perform clustering
```{r}
repl <- 1:5
gens <- c(0,10,20,30,40,50,60)

#fluctutaing
cands_fluc <- readRDS("Fluc_F0_F60_cmh_sig_AF.rds")
cmh_fluc <- readRDS("Fluc_cmh_p.rds")

#get haplovalidate parameters
parameters <- get.mncs.win(cands_fluc, cmh_fluc, wins = seq(0.1, 10, 0.05), mncs = 0.01)

#run haplovalidate
happy_fluc <- haplovalidate(cands_fluc, 
                       cmh_fluc, 
                       parameters,
                       repl,
                       gens,
                       takerandom = 2000,
                       filterrange = 5000)
saveRDS(happy_fluc,file="Fluc_F0_F60_haplotypes.rds")


##constant
cands_cons <- readRDS("Cons_F0_F60_cmh_sig_AF.rds")
cmh_cons <- readRDS("cons_cmh_p.rds")

#get haplovalidate parameters
parameters <- get.mncs.win(cands_cons, cmh_cons, wins = seq(0.1, 10, 0.05), mncs = 0.01)

#run haplovalidate
happy_cons <- haplovalidate(cands_cons, 
                       cmh_cons, 
                       parameters,
                       repl,
                       gens,
                       takerandom = 2000,
                       filterrange = 5000)
saveRDS(happy_cons,file="Cons_F0_F60_haplotypes.rds")
```


#3. visualisation
###3.1 modified function of plot
```{r}
plot.haplovalidate.new <- function(blocks,cmh,title="",label=TRUE){
     neuefarbe.raw <- unique(c(brewer.pal(8,"Set1"),brewer.pal(7,"Dark2"),brewer.pal(7,"Set2"),brewer.pal(12,"Set3"),brewer.pal(12,"Paired"),brewer.pal(7,"Accent"),brewer.pal(7,"Spectral")))
     neuefarbe  <- colorRampPalette(neuefarbe.raw)(max(as.numeric(factor(blocks$tag)),na.rm=TRUE))
    ## as.numeric removes the non-grouped tags!
    col.sub <- neuefarbe[as.numeric(factor(blocks$tag))]
    blocks[,groupcol := col.sub]
        for (j in unique(blocks$chr)){
            cluster.sub <- merge(blocks[chr==j,],cmh,by=c("chr","pos"))
            png(paste("Plot_",title,"_",j,"_haplovalidate.png",sep=""),2400,1200)
            par(mar = c(12,10,6,2))
            maxy <- round(max(as.numeric(cmh[chr==j,score]),na.rm=TRUE),0) + round(max(as.numeric(cmh[chr==j,score]),na.rm=TRUE),0) * 0.02
            cmh[chr==j,plot(pos/1000000,as.numeric(score),pch=19,col="#414547",ylim=c(0,maxy),
                            xaxt = "n", yaxt ="n",
                            xlab="",ylab="",lwd=3,
                            #cex.axis=4,cex.lab=6,
                            #main=paste0(title,"_chr",j), adj = 0, cex.main = 10, 
                            xaxp=c(0,50,10))]
            axis(1, mgp = c(3,3,0), cex.axis = 4) #xaxis
            axis(2, mgp = c(3,2,0), cex.axis = 4) #yaxis
            mtext("position Mb", side = 1, line = 8, cex = 6) #xlab
            mtext("cmh score", side = 2, line = 6, cex = 6) #ylab
            title(main = paste0(title,"_chr",j), adj = 0, cex.main = 6)
            cluster.sub[,points(as.numeric(pos)/1000000,as.numeric(score),cex=2,lwd=3,pch=19,col=groupcol)]
            count <- 0
            cluster.sub.ord <- cluster.sub[order(score,decreasing=TRUE),]
            for (k in na.exclude(unique(cluster.sub.ord$groupcol))){
                group.sub <- cluster.sub.ord[groupcol==k,]
                distance <- group.sub$pos/1000000
                levi <- maxy  - count 
                points(distance,rep(levi,length(distance)),col=k,pch=19,cex=3)
                if (label)
                    text(distance[1],levi,unique(group.sub[,tag]),cex =2)
                count <- count +  round(max(as.numeric(cmh[chr==j,score]),na.rm=TRUE),0) * 0.01
            }
            dev.off()
        }
     return(blocks)
 }

```

###3.2 plot haplotype
```{r}
happy_fluc <- readRDS("Fluc_F0_F60_haplotypes.rds")
cmh <- readRDS("Fluc_cmh_p.rds")

plot.haplovalidate.new(blocks=happy_fluc$dominant_haplotypes,cmh,title="Fluc",label=T)

happy_cons <- readRDS("Cons_F0_F60_haplotypes.rds")
cmh <- readRDS("Cons_cmh_p.rds")

plot.haplovalidate.new(blocks=happy_cons$dominant_haplotypes,cmh,title="Cons",label=T)
```

#4. check the overlapping of haplotypes across treatment

###4.1 find the haplotypes that overlap
```{r}
#setwd("~/my_project/00upload_github/genome_biology_florida_fluc_cons/07revision_v2/haplotype_analysis/")

#load the haplovalidate results
happy_fluc<-readRDS("Fluc_F0_F60_haplotypes.rds")
happy_cons<-readRDS("Cons_F0_F60_haplotypes.rds")


##fluctuating
haplo_info_fluc<- happy_fluc$dominant_haplotypes
af_fluc <- readRDS("Fluc_F0_F60_cmh_sig_AF.rds")
haplo_info_fluc[af_fluc, on = .(chr,pos), riseallele := i.riseallele]
#haplo_fluc <- haplo_info_fluc[, .(
#  start = min(pos),
#  end = max(pos)
#  ), by = .(chr, tag)]

#constant
haplo_info_cons<- happy_cons$dominant_haplotypes
af_cons <- readRDS("Cons_F0_F60_cmh_sig_AF.rds")
haplo_info_cons[af_cons, on = .(chr,pos), riseallele := i.riseallele]

#haplo_cons <- haplo_info_cons[, .(
#  start = min(pos),
#  end = max(pos)
#), by = .(chr, tag)]

#check the overlapping between haplotypes across environments
overlap <- list()
i <- 1
for (haplo_id_fluc in unique(haplo_info_fluc$tag)) {
  for (haplo_id_cons in unique(haplo_info_cons$tag)){
    #information for the haplotype in the fluc
    fluc_chr <- unique(haplo_info_fluc[tag==haplo_id_fluc,chr])
    fluc_start <- min(haplo_info_fluc[tag==haplo_id_fluc,pos])
    fluc_end <- max(haplo_info_fluc[tag==haplo_id_fluc,pos])
    
    #information for the haplotype in the cons
    cons_chr <- unique(haplo_info_cons[tag==haplo_id_cons,chr])
    cons_start <- min(haplo_info_cons[tag==haplo_id_cons,pos])
    cons_end <- max(haplo_info_cons[tag==haplo_id_cons,pos])
    
    if (fluc_chr==cons_chr & max(fluc_start,cons_start) < min(fluc_end,cons_end)){ #check if the two haplotype overlaps
      #overlapwindow
      inter_chr <- fluc_chr;inter_start <- max(fluc_start,cons_start);inter_end <- min(fluc_end,cons_end)
      #getting snps from the overlapping window 
      fluc_snp <- haplo_info_fluc[tag == haplo_id_fluc & chr == inter_chr & pos >= inter_start & pos <= inter_end, .(chr,pos,riseallele)]
      cons_snp <- haplo_info_cons[tag == haplo_id_cons & chr == inter_chr & pos >= inter_start & pos <= inter_end, .(chr,pos,riseallele)]
      #check how many snps are overlapping
      num_overlap <- nrow(fluc_snp[cons_snp, on = .(chr,pos,riseallele), nomatch = 0L])
      num_total <- nrow(fluc_snp) + nrow(cons_snp) - num_overlap
      percent_fluc <- num_overlap/nrow(fluc_snp)
      percent_cons <- num_overlap/nrow(cons_snp)
      percent_overlap <- num_overlap/num_total
      
      #write output
      tmp <- data.table(
        haplotype_id_fluc <- haplo_id_fluc,
        fluc_chr <- fluc_chr,
        fluc_start <- fluc_start,
        fluc_end <- fluc_end,
        haplotype_id_con <- haplo_id_cons,
        cons_chr <- cons_chr,
        cons_start <- cons_start,
        cons_end <- cons_end,
        intersect_chr <- inter_chr,
        intersect_start <- inter_start,
        intersect_end <- inter_end,
        num_snp_fluc <- nrow(fluc_snp),
        num_snp_cons <- nrow(cons_snp),
        num_snp_total <- num_total,
        num_snp_shared <- num_overlap,
        percent_fluc <- percent_fluc,
        percent_cons <- percent_cons,
        percent_share <- percent_overlap
      )
      overlap[[i]] <- tmp
      
      i <- i + 1
    }
  }
}

dt_overlap <- rbindlist(overlap)
setnames(dt_overlap,c("haplotype_id_fluc","fluc_chr","fluc_start","fluc_end",
                      "haplotype_id_cons","cons_chr","cons_start","cons_end",
                      "intersect_chr","intersect_start","intersect_end",
                      "num_snp_in_fluc","num_snps_in_cons","num_snps_total","num_snps_shared",
                      "percent_in_fluc","percent_in_cons","percent_share"))
saveRDS(dt_overlap, file = "Haplotype_sharing.rds")

```

###4.2 summary plot
```{r}
dt_overlap <- readRDS("Haplotype_sharing.rds")
p_overlap <- ggplot(dt_overlap, aes(x=percent_share*100))+
  geom_density()+
  labs(x="%SNP shared")+
  theme_bw()+
  theme(axis.text = element_text(size =10),
        axis.title = element_text(size=12))
ggsave("check_overlapping/summary_percent_share.pdf", plot = p_overlap, width = 10, height = 5)

```

###4.3 top 5 haplo-pair with higheest sharing
```{r}

haplo_info_fluc<- readRDS("Fluc_F0_F60_haplotypes.rds")$dominant_haplotypes
haplo_info_cons<- readRDS("Cons_F0_F60_haplotypes.rds")$dominant_haplotypes

plot_hap_pair <- function(tag_fluc, tag_cons, label_id){
  col_shared <- "#6e326e"
  col_cons <- "#4fc0c6"
  col_fluc <- "#e8a11c"
    
  hap_fluc_snp<-haplo_info_fluc[tag==tag_fluc,]
  hap_cons_snp<-haplo_info_cons[tag==tag_cons,]

  shared_snp <- merge(
    hap_fluc_snp[, .(chr,pos)],
    hap_cons_snp[, .(chr,pos)]
  )[,pos]

  invisible(hap_fluc_snp[, category := fifelse(pos %in% shared_snp, "shared", "fluc")])
  invisible(hap_cons_snp[, category := fifelse(pos %in% shared_snp, "shared", "cons")])

  #get window
  fluc_start<-min(hap_fluc_snp$pos); fluc_end<-max(hap_fluc_snp$pos)
  cons_start<-min(hap_cons_snp$pos);cons_end<-max(hap_cons_snp$pos)
  overlap_start <- max(fluc_start, cons_start); overlap_end <- min(fluc_end, cons_end)

  #overlapping haplotypes
  p1<-ggplot()+
    #plot the haplotype
    annotate("rect",xmin = fluc_start, xmax = fluc_end, ymin =0.9, ymax=1.1, fill = NA, color=col_fluc)+
    annotate("text", x = fluc_start, y = 1.3, hjust=-0.1, ,label = tag_fluc, color=col_fluc)+
    annotate("rect", xmin = cons_start, xmax = cons_end, ymin =1.9, ymax=2.1, fill=NA,color=col_cons)+
    annotate("text", x = cons_end, y = 2.3, hjust=1.1, ,label = tag_cons, color=col_cons)+
    #add significant SNPs on the haplotype
    geom_point(data = hap_fluc_snp, aes(x=pos,y=1, color=category), size = 0.6)+
    geom_point(data = hap_cons_snp, aes(x=pos,y=2, color=category), size = 0.6)+
    scale_color_manual(values = c(col_cons, col_fluc, col_shared))+
    #highlight overlaaping region
    geom_rect(aes(xmin = overlap_start, xmax = overlap_end, ymin = 0.5, ymax = 2.5), fill = NA, linetype="dashed", color="black")+
    #xlim(overlap_start-1000,overlap_end+1000)+
    ylim(0,3)+
    labs(x="position",y="")+
    theme_minimal()+
    theme(axis.text.y = element_blank(),
          axis.text.x = element_text(size = 10),
          axis.title = element_text(size = 10),
          plot.margin = margin(t = 5, r = 20, b = 5, l = 5))
  #plot(p1)
  
  #zoom in to the overlap region
  p2<-ggplot()+
    #plot the haplotype
    annotate("rect",xmin = overlap_start, xmax = overlap_end, ymin =0.9, ymax=1.1, fill = NA, color=col_fluc)+
    annotate("text", x = overlap_start, y = 1.3, hjust=-0.1, ,label = tag_fluc, color=col_fluc)+
    annotate("rect", xmin = overlap_start, xmax = overlap_end, ymin =1.9, ymax=2.1, fill=NA,color=col_cons)+
    annotate("text", x = overlap_end, y = 2.3, hjust=1.1, ,label = tag_cons, color=col_cons)+
    #add significant SNPs on the haplotype
    geom_point(data = hap_fluc_snp, aes(x=pos,y=1, color=category), size = 0.6)+
    geom_point(data = hap_cons_snp, aes(x=pos,y=2, color=category), size = 0.6)+
    scale_color_manual(values = c(col_cons, col_fluc, col_shared))+
    #highlight overlaaping region
    geom_rect(aes(xmin = overlap_start, xmax = overlap_end, ymin = 0.5, ymax = 2.5), fill = NA, linetype="dashed", color="black")+
    xlim(overlap_start,overlap_end)+
    ylim(0,3)+
    labs(x="position",y="")+
    theme_minimal()+
    theme(axis.text.y = element_blank(),
          axis.text.x = element_text(size = 10),
          axis.title = element_text(size = 10),
          plot.margin = margin(t = 5, r = 20, b = 5, l = 5))
  #plot(p2)

  merge_p<-ggarrange(p1,p2,nrow=2, common.legend = TRUE)
  merge_p<-annotate_figure(p = merge_p, top= text_grob(label_id, x = 0.05, hjust = 0, face = "bold", size = 14))
  ggsave(paste0("check_overlapping/",label_id,".pdf"),plot = merge_p, width = 10, height = 5)
  #plot(merge_p)  
}

sorted_dt_overlap<-dt_overlap[order(-percent_share)]

#check the top 5 overlapping pairs(>30% shared)

for (i in 1:5){
  tag_fluc<-sorted_dt_overlap[i, haplotype_id_fluc]
  tag_cons<-sorted_dt_overlap[i, haplotype_id_cons]
  percent_share <- sorted_dt_overlap[i, percent_share]
  label_id <- paste0("Fluc(",tag_fluc, ")_Cons(",tag_cons,")_",round(percent_share,4))
  plot_hap_pair(tag_fluc, tag_cons, label_id)
}

```

##4.4 count region overlap
```{r}
##function to get the haplotype regions
get_bed <- function(haplo_info){
  haplo_bed <- list()
  i=1
  for (haplo_id in unique(haplo_info$tag)){
    chr <- unique(haplo_info[tag==haplo_id,chr])
    start <- min(haplo_info[tag==haplo_id,pos])
    end <- max(haplo_info[tag==haplo_id,pos])
    haplo_bed[[i]] <- data.table(haplo_id=haplo_id, chr = chr, start = start, end = end)
    i <- i + 1
  }
  haplo_bed<-rbindlist(haplo_bed)
  return(haplo_bed)
}

merge_overlap <- function(haplo_bed){
  setorder(haplo_bed, chr, start, end)
  
  haplo_bed[, max_end := cummax(end), by = chr]
  haplo_bed[, group := cumsum(start > shift(max_end, fill = first(start) -1)), by = chr]
  
  merged_bed <- haplo_bed[, .(
    start = min(start),
    end = max(end)
  ), by = .(chr, group)]
   
  merged_bed[, group := NULL]
  merged_bed[, length := end - start +1 ]
  
  return(merged_bed)
}

#fluctuating
haplo_info_fluc<-readRDS("Fluc_F0_F60_haplotypes.rds")$dominant_haplotypes
haplo_fluc<-get_bed(haplo_info_fluc)
merged_haplo_fluc<-merge_overlap(haplo_fluc)
len_fluc<-sum(merged_haplo_fluc[, length])
#constant
haplo_info_cons<-readRDS("Cons_F0_F60_haplotypes.rds")$dominant_haplotypes
haplo_cons<-get_bed(haplo_info_cons)
merged_haplo_cons<-merge_overlap(haplo_cons)
len_cons<-sum(merged_haplo_cons[, length])

#get the total length
#combine bed file from fluc and cons
combine_haplo <- merge_overlap(rbind(merged_haplo_fluc, merged_haplo_cons))
len_total <- sum(combine_haplo[,length])

#check the overlap
hap_pair_overlap <- readRDS("Haplotype_sharing.rds")
overlap_bed <- data.table(chr = hap_pair_overlap$intersect_chr,
                          start = hap_pair_overlap$intersect_start,
                          end = hap_pair_overlap$intersect_end)
merged_overlap <- merge_overlap(overlap_bed)
len_overlap <- sum(merged_overlap[,length])

exp_overlap<-len_cons*len_fluc/len_total
```
#5.check the shared SNPs
crosscheck with the CMH(F60) results, check whether the shared SNPs enriched in the haplotype


```{r}
#update the position 2L, 2R --> 2 3L,3R -->3
len_2L=as.integer(21106324)
len_3L=as.integer(22263419)

#load cmh results
fluc_cmh <- fread("../00data/genomic_data/fluctuating_F0_F60_CMH.result.gz", header = TRUE)
setnames(fluc_cmh, old = c("p", "adj_p"), new = c("fluc_p", "fluc_adj_p"))



cons_cmh <- fread("../00data/genomic_data/constant_F0_F60_CMH.result.gz", header=TRUE)
setnames(cons_cmh, old = c("p", "adj_p"), new = c("cons_p", "cons_adj_p"))


cmh <- merge(fluc_cmh, cons_cmh, by = c("chr","pos"), all = TRUE)
cmh[is.na(cmh)] <- 1
#keep only the significant SNPs
cmh_sig <- cmh[fluc_adj_p < 0.05 | cons_adj_p<0.05,]
#add category based on the significance
cmh_sig[, significance := fifelse(fluc_adj_p<0.05 & cons_adj_p < 0.05, "Shared",
                                  fifelse(fluc_adj_p<0.05, "Fluctuating", "Constant"))]
cmh_sig[, significance := factor(significance, levels = c("Shared", "Constant", "Fluctuating"))]
cmh_sig[, loci := paste0(chr, ".", pos)]

##load haplotype information
fluc_hap <- readRDS("Fluc_F0_F60_haplotypes.rds")$dominant_haplotypes
af_fluc <- readRDS("Fluc_F0_F60_cmh_sig_AF.rds")
fluc_hap[af_fluc, on = .(chr,pos), riseallele := i.riseallele]

cons_hap <- readRDS("Cons_F0_F60_haplotypes.rds")$dominant_haplotypes
af_cons <- readRDS("Cons_F0_F60_cmh_sig_AF.rds")
cons_hap[af_cons, on = .(chr,pos), riseallele := i.riseallele]

#hap_pair between fluc and cons
hap_pair_overlap <- readRDS("Haplotype_sharing.rds")
```
##functions
```{r}
#manhattn plot
plot_manhattan <- function(dt, y_col, plot_title){

  p_manhattan<-ggplot(dt, aes(x=pos/1000000, y=-log10({{ y_col }}), color = significance))+
    geom_point(cex=0.1)+
    theme(panel.grid = element_blank(), panel.background = element_rect(fill="white", color= "black"),
          axis.text.x = element_text(size=8),axis.text.y = element_text(size=10),
          axis.title = element_text(size=12), title = element_text(size=12),legend.position = "none")+
    #scale_x_continuous(breaks = seq(0,25,5))+
    scale_y_continuous(breaks = seq(0,20,5), limits = c(0,20))+
    geom_hline(aes(yintercept=-log10(0.05)), color="red", linetype="dashed")+
    scale_color_manual(values = c("#6e326e","#4fc0c6","#e8a11c"))+
    labs(x = "postion(Mb)", y = "CMH -log10(p-value)", title=plot_title)

  return(p_manhattan)

}


#haplotype_plot
plot_haplotype <- function(chrom, start, end){

  len_2L=as.integer(21106324)
  len_3L=as.integer(22263419)

  cols <- c(
    Fluctuating = "#e8a11c",
    Constant = col_cons <- "#4fc0c6",
    Shared = "#6e326e", #SNPs significant in both pop and same allele increase freq
    Divergence = "grey" #SNPs significant in both pop but different allele increase freq
  )
  col_levels <- c("Divergence", "Shared", "Fluctuating", "Constant")
  
  hap_chr <- fifelse(
    chrom %in% c("2L","2R"), "2",
    fifelse(chrom %in% c("3L", "3R"), "3", "X")
    )
  hap_start <- fifelse(
    chrom == "2R", start + len_2L,
    fifelse(chrom == "3R", start + len_3L, start)
  )
  hap_end <- fifelse(
    chrom == "2R", end + len_2L,
    fifelse(chrom == "3R", end + len_3L, end)
  )

  #fluc haplotype in this region
  dt_hap <- fluc_hap[chr == hap_chr & pos > hap_start & pos < hap_end,]
  dt_hap[, cmh_chr := fifelse(chr=="2", fifelse(pos < len_2L, "2L", "2R"),
                              fifelse(chr == "3", fifelse(pos < len_3L, "3L", "3R"), "X")
                              )]
  dt_hap[, cmh_pos := fifelse(cmh_chr== "2R", pos - len_2L,
                              fifelse(cmh_chr =="3R", pos - len_3L, pos))]  
  dt_hap[, loci := paste0(cmh_chr, ".",cmh_pos)]

  dt_hap<-merge(dt_hap, cmh_sig[, c("loci","fluc_adj_p", "significance")], by = "loci")
  
  #update the significance by checking if the same alleles increase freq in fluc and cons
  cons_shared <- af_cons[, .(chr, pos, cons_rise_allele = riseallele)]
  dt_hap[cons_shared, cons_rise_allele := i.cons_rise_allele, on = c("chr","pos")]
  dt_hap[, new_significance := fifelse(significance=="Fluctuating","Fluctuating",
                                       fifelse(riseallele==cons_rise_allele, "Shared", "Divergence"))]
  dt_hap[, new_significance := factor(new_significance, levels = col_levels)]
  
  hap_summary <- dt_hap[, .(
    n_snps = uniqueN(cmh_pos),
    n_shared_snps = uniqueN(cmh_pos[new_significance == "Shared"]),
    start = min(cmh_pos),
    end = max(cmh_pos)
  ),
  by = tag]

  hap_max <- hap_pair_overlap[, .(max_percent_sharing = max(percent_share, na.rm = TRUE)), by = haplotype_id_fluc]
  hap_max[, max_percent_sharing := sprintf("%.2f%%", 100*max_percent_sharing)]
  hap_summary[hap_max, max_percent_sharing := i.max_percent_sharing, on = .(tag = haplotype_id_fluc) ]
  
  p_fluc <- ggplot(dt_hap, aes(x=cmh_pos/1000000, y =tag, color = new_significance))+
    geom_point(cex=0.1)+
    geom_text(data = hap_summary, 
              aes(x=end/1000000, y=tag, label = max_percent_sharing), 
              inherit.aes = FALSE, vjust = 1.5, hjust = 1)+
    xlim(start/1000000-0.5,end/1000000+0.5)+
    scale_color_manual(values = cols)+
    labs(x = "postion(Mb)", 
         y = "Haplotype ID", 
         title="Fluctuating Haplotypes")+
    theme(panel.grid = element_blank(), panel.background = element_rect(fill="white", color= "black"),
          axis.text.x = element_text(size=8),axis.text.y = element_text(size=10),
          axis.title = element_text(size=12), title = element_text(size=12),
          legend.position = "none")

  #plot(p_fluc)
  
  #cons haplotype in this region
  dt_hap <- cons_hap[chr == hap_chr & pos > hap_start & pos < hap_end,]
  dt_hap[, cmh_chr := fifelse(chr=="2", fifelse(pos < len_2L, "2L", "2R"),
                              fifelse(chr == "3", fifelse(pos < len_3L, "3L", "3R"), "X")
                              )]
  dt_hap[, cmh_pos := fifelse(cmh_chr== "2R", pos - len_2L,
                              fifelse(cmh_chr =="3R", pos - len_3L, pos))]  
  dt_hap[, loci := paste0(cmh_chr, ".",cmh_pos)]

  dt_hap<-merge(dt_hap, cmh_sig[, c("loci","fluc_adj_p", "significance")], by = "loci")
  
  #update the significance by checking if the same alleles increase freq in fluc and cons
  fluc_shared <- af_fluc[, .(chr, pos, fluc_rise_allele = riseallele)]
  dt_hap[fluc_shared, fluc_rise_allele := i.fluc_rise_allele, on = c("chr","pos")]
  dt_hap[, new_significance := fifelse(significance=="Constant","Constant",
                                       fifelse(riseallele==fluc_rise_allele, "Shared", "Divergence"))]
  dt_hap[, new_significance := factor(new_significance, levels = col_levels)]
  
  hap_summary <- dt_hap[, .(
    n_snps = uniqueN(cmh_pos),
    n_shared_snps = uniqueN(cmh_pos[new_significance == "Shared"]),
    start = min(cmh_pos),
    end = max(cmh_pos)
  ),
  by = tag]

  #hap_summary[, percent_share := sprintf("%.2f%%", 100 * n_shared_snps/n_snps)]
  hap_max <- hap_pair_overlap[, .(max_percent_sharing = max(percent_share, na.rm = TRUE)), by = haplotype_id_cons]
  hap_max[, max_percent_sharing := sprintf("%.2f%%", 100*max_percent_sharing)]
  hap_summary[hap_max, max_percent_sharing := i.max_percent_sharing, on = .(tag = haplotype_id_cons) ]
  
  p_cons <- ggplot(dt_hap, aes(x=cmh_pos/1000000, y =tag, color = significance))+
    geom_point(cex=0.1)+
    geom_text(data = hap_summary, 
              aes(x=end/1000000, y=tag, label = max_percent_sharing), 
              inherit.aes = FALSE, vjust = 1.5, hjust = 1)+
    xlim(start/1000000-0.5,end/1000000+0.5)+
    scale_color_manual(values = cols)+
    labs(x = "postion(Mb)", 
         y = "Haplotype ID", 
         title="Constant haplotypes")+
    theme(#panel.grid = element_blank(), 
          panel.background = element_rect(fill="white", color= "black"),
          axis.text.x = element_text(size=8),axis.text.y = element_text(size=10),
          axis.title = element_text(size=12), title = element_text(size=12),
          legend.position = "none")


  p_haplotype <- ggarrange(p_fluc, p_cons, nrow = 2)
  return(p_haplotype)
}
```


##5.1 chr2R_10Mb(7.5-12.5)
```{r}
chrom = "2R"; pos = 10000000; start = pos - 2500000; end = pos + 2500000
dt<-cmh_sig[chr==chrom & pos > start & pos < end,]
p_fluc <- plot_manhattan(dt, fluc_adj_p, plot_title = "Fluc vs Anc")
p_cons <- plot_manhattan(dt, cons_adj_p, plot_title = "Cons vs Anc")
p_manhattan <- ggarrange(p_fluc, p_cons, nrow = 2)

#haplotype plot
p_haplotype <- plot_haplotype(chrom, start, end)

p <- ggarrange(p_manhattan, p_haplotype, ncol = 2)
p <- annotate_figure(p, 
                     top = text_grob(
                       paste0("Chr",chrom,"_",pos/1000000, "Mb (",
                              start/1000000, "-", end/1000000, "Mb)"),
                       x = 0, hjust = 0, face = "bold", size = 16))

ggsave(filename = paste0("zoom_in_check/",chrom,"_",start,"_",end,".pdf"),
      plot = p, width = 10, height = 8)
```

##5.2 chr3R_5Mb(2.5-7.5)
```{r}
chrom = "3R"; pos = 5000000; start = pos - 2500000; end = pos + 2500000
dt<-cmh_sig[chr==chrom & pos > start & pos < end,]
p_fluc <- plot_manhattan(dt, fluc_adj_p, plot_title = "Fluc vs Anc")
p_cons <- plot_manhattan(dt, cons_adj_p, plot_title = "Cons vs Anc")
p_manhattan <- ggarrange(p_fluc, p_cons, nrow = 2)

#haplotype plot
p_haplotype <- plot_haplotype(chrom, start, end)

p <- ggarrange(p_manhattan, p_haplotype, ncol = 2)
p <- annotate_figure(p, 
                     top = text_grob(
                       paste0("Chr",chrom,"_",pos/1000000, "Mb (",
                              start/1000000, "-", end/1000000, "Mb)"),
                       x = 0, hjust = 0, face = "bold", size = 16))

ggsave(filename = paste0("zoom_in_check/",chrom,"_",start,"_",end,".pdf"),
      plot = p, width = 10, height = 8)
```

##5.2 chrX_10Mb(2.5-7.5)
```{r}
chrom = "X"; pos = 10000000; start = pos - 2500000; end = pos + 2500000
dt<-cmh_sig[chr==chrom & pos > start & pos < end,]
p_fluc <- plot_manhattan(dt, fluc_adj_p, plot_title = "Fluc vs Anc")
p_cons <- plot_manhattan(dt, cons_adj_p, plot_title = "Cons vs Anc")
p_manhattan <- ggarrange(p_fluc, p_cons, nrow = 2)

#haplotype plot
p_haplotype <- plot_haplotype(chrom, start, end)

p <- ggarrange(p_manhattan, p_haplotype, ncol = 2)
p <- annotate_figure(p, 
                     top = text_grob(
                       paste0("Chr",chrom,"_",pos/1000000, "Mb (",
                              start/1000000, "-", end/1000000, "Mb)"),
                       x = 0, hjust = 0, face = "bold", size = 16))

ggsave(filename = paste0("zoom_in_check/",chrom,"_",start,"_",end,".pdf"),
      plot = p, width = 10, height = 8)
```



```{r}
hap_overlap<-readRDS("Haplotype_sharing.rds")
```