---
title: "competitive_fitness_bothCGE"
author: "Changyi"
date: "2026-02-18"
output: html_document
---



```{r}
library(R.utils)
library(pastecs)
library(data.table)
library(stringr)
library(ggsignif)
library(lme4)
library(emmeans)
library(ggplot2)
library(readxl)
library(Rmisc)
library(readxl)
library(ggpubr)
library(readxl)
library(Rmisc)
```

#1.competitive fitness
##1.1 load data
```{r}
load_data <- function(CGE=NULL) {
  pop <- paste0(rep(c("Anc_R", "Constant_R", "wave2_R"), each = 5), rep(1:5, 3))
  
  raw_data <- read_xlsx(paste0("competitivefitness/",CGE,"_competitivefitness_rawdata.xlsx"))
  raw_data <- subset(raw_data, EvoRep %in% pop)
  #get accumulative results
  #accumu_data<-NULL
  raw_data <- subset(raw_data, day %in% 1:10)
  t1 <- aggregate(raw_data$corrected_white,
                  by = list(raw_data$EvoRep, raw_data$TechRep),
                  sum)
  t2 <- aggregate(raw_data$corrected_red,
                  by = list(raw_data$EvoRep, raw_data$TechRep),
                  sum)
  accumu_data <- data.table(
    evorep = t1$Group.1,
    techrep = t1$Group.2,
    white = t1$x,
    red = t2$x,
    Days = 10,
    Population = rep(rep(
      c("Ancestral", "Constant", "Fluctuating"), each = 5
    ), 3)
  )
  #accumu_data <- rbind(accumu_data,data1)
  invisible(accumu_data[, total := red + white])
  invisible(accumu_data[, ratio := red / total])
  invisible(accumu_data[, CGE := fifelse(CGE=="28_18", "2818", "23")])
  return(accumu_data)
}

data_2818<-load_data(CGE="28_18")
data_23<-load_data(CGE="23")
data_merge<-rbind(data_23,data_2818)
invisible(data_merge[, Population := factor(Population, levels = c("Fluctuating","Constant","Ancestral"))]) #we modify the order of levels to keep the sign positive
invisible(data_merge[, over := factor(rownames(data_merge))]) #to account for the overdispersion
cols_to_factor<-c("evorep","techrep","Population","CGE")
invisible(data_merge[, (cols_to_factor) := lapply(.SD, factor), .SDcols = cols_to_factor])
rm(data_23, data_2818, cols_to_factor)
```


##1.2 fit data to the model(glmer)
```{r}
#full_model<-glmer(ratio ~ CGE*Population + (1|evorep), data=data_merge, family = binomial, weights = total)
full_model<-glmer(ratio ~ CGE*Population + (1|evorep) + (1|over), data=data_merge, family = binomial, weights = total)
simple_model<-glmer(ratio ~ Population + (1|evorep) + (1|over), data=data_merge, family = binomial, weights = total)

```

##1.3 check the overdispersion factor
```{r}
source('competitivefitness/diagnostic_fcns.r')
diag_full<-overdisp.test(full_model)
diag_full
diag_simple<-overdisp.test(simple_model)
diag_simple
rm(diag_full, diag_simple)
```

##1.4 model competing
```{r}
anova(full_model, simple_model)
rm(simple_model)
```

##1.5 model output
```{r}
summary(full_model)
```

##1.6 emmeans
```{r}
relfit_emmeans<-emmeans(full_model, pairwise ~ Population|CGE, adjust="none", type="link")

relfit_emmeans


```
##1.7 compare the contrast
```{r}
#this code asking whether the differences between anc and evolved population are significantly larger than the differences between fluc and cons populations

emm<-summary(relfit_emmeans)
#use differences between fluc and cons as the null model to test against
obs_diff<-emm$emmeans[emm$emmeans$Population=="Fluctuating","emmean"]-
  emm$emmeans[emm$emmeans$Population=="Constant","emmean"]

##the population ordered as fluc, cons, anc
#test whether (fluc-anc) larger than (fluc-cons)
#(fluc-anc)-(fluc-cons)=0*fluc+1*cons+(-1)*anc
contrast(relfit_emmeans, list(comparison = c(0,1,-1)), by="CGE", side = ">")
#contrast(relfit_emmeans, list(comparison = c(1, 0, -1)), by ="CGE", null = obs_diff) #p<0.0001 alternative contrast that follow the same logic

#test whether (cons-anc) larger than (fluc-cons)
#(cons-anc)-(fluc-cons)=(-1)*fluc+2*cons+(-1)*anc. (1,2,-1)
contrast(relfit_emmeans, list(fc_vs_fa = c(1,2,-1)), by="CGE", side = ">")
#contrast(relfit_emmeans,list(comparison = c(0, 1, -1)), by ="CGE", null = obs_diff) #p<0.0001 alternative contrast that follow the same logic

```


#2.fecundity


##2.1 load fecundity data
```{r}
##load fecundity raw data
eggs_alldates_2818 <- fread("fecundity/rawdata_2818cge.csv")
eggs_alldates_2818 <- eggs_alldates_2818[pop %in% c("anc","constant","wave2")]
invisible(eggs_alldates_2818[, CGE := 2818])
eggs_alldates_23 <- fread("fecundity/rawdata_23cge.csv")
eggs_alldates_23 <- eggs_alldates_23[pop %in% c("anc","constant","wave2")]
invisible(eggs_alldates_23[pop %in% c("anc","constant","wave2"), CGE := 23])
eggs_alldates<-rbind(eggs_alldates_23, eggs_alldates_2818)
#sum up the eggs across days
eggs_sum <- eggs_alldates[, .(n_egg_sum = sum(n_egg)), by = .(pop, rep, run, CGE)]
invisible(eggs_sum[, Population := fifelse(pop=="anc","Ancestral",
                                 fifelse(pop=="constant","Constant","Fluctuating"))])
invisible(eggs_sum[, sample := paste(pop, run, rep, sep="_")])
#load parents information
info_fluc<-read_xlsx("fecundity/fluc_fecundity_summary.xlsx", col_names = TRUE);info_fluc$CGE <- 2818
info_cons<-read_xlsx("fecundity/cons_fecundity_summary.xlsx", col_names = TRUE);info_cons$CGE <- 23
info<-rbind(info_fluc,info_cons)
merged_dt<-merge(eggs_sum,info, by=c("sample","CGE"))
#only keep the needed information
merged_dt<-merged_dt[, .(sample,Population, rep, run, n_egg_sum, number_of_female, weight_females, CGE)]

colnames(merged_dt)<-c("sample","Population","rep","run","num_egg","number_of_female","weight_females","CGE")
invisible(merged_dt[, evorep := paste0(Population, "_", rep)])
cols_to_factor <- c("sample","Population","rep","run","evorep","CGE")
merged_dt[, (cols_to_factor) := lapply(.SD, factor), .SDcols = cols_to_factor]
rm(eggs_alldates, eggs_alldates_23, eggs_alldates_2818, eggs_sum, info, info_cons, info_fluc, cols_to_factor)
```




##2.2 lmer

```{r}
fecundity_lmer<-lmer(log10(num_egg/number_of_female) ~ CGE*Population + (1|evorep), data=merged_dt, REML = FALSE)
summary(fecundity_lmer)
```





##2.3 if residuals are normally distributed
```{r}
ks.test(resid(fecundity_lmer),"pnorm", mean(resid(fecundity_lmer)), sd(resid(fecundity_lmer))) #pvalue=0.9835
qqnorm(resid(fecundity_lmer))
```
##2.4 check the homogeneity of variance
```{r}
plot(fecundity_lmer) #if residual evenly distributed around the center line

```
##2.5 emmeans
```{r}
fecunity_emmeans<-emmeans(fecundity_lmer, pairwise ~ Population |CGE, adjust="none")
fecunity_emmeans
```
#3. plot the phenotyic data

##3.1 multiple testing correction
```{r}
#here we perform multiple testing correction jointly for competitive fitness and fecundity, as the same question being asked
#contrast
#competitive fitness
contrast_competitivefitness <- as.data.table(relfit_emmeans$contrasts)
contrast_competitivefitness$trait <- "Competitive Fitness"
#fecundity
contrast_fecundity <- as.data.table(fecunity_emmeans$contrasts)
contrast_fecundity$trait <- "Fecundity"

contrast_result <- rbindlist(list(contrast_fecundity, contrast_competitivefitness), fill = TRUE)
invisible(contrast_result[, adj.p := p.adjust(p.value)])
contrast_result[, significant := fifelse(adj.p>0.05, "ns",
                                         fifelse(adj.p>0.01, "*",
                                                 fifelse(adj.p>0.001, "**",
                                                         fifelse(adj.p>0.0001,"***","****"))))]
#emmeans

##competitive fitness
relfit_emmeans1<-emmeans(full_model, pairwise ~ Population|CGE, adjust="none", type="response")
emmeans_competitivefitness <- as.data.table(relfit_emmeans1$emmeans)
emmeans_competitivefitness$trait <- "Competitive Fitness"
##fecundity
fecunity_emmeans1<-emmeans(fecundity_lmer, pairwise ~ Population |CGE, adjust="none", type="response")
emmeans_fecundity <- as.data.table(fecunity_emmeans1$emmeans)
emmeans_fecundity$trait <- "Fecundity"

emmeans_result <- rbindlist(list(emmeans_fecundity, emmeans_competitivefitness), fill = TRUE)
emmeans_result$Population<-factor(emmeans_result$Population, levels = c("Ancestral", "Constant", "Fluctuating"))
```


##3.2 plot from model fit
```{r}
p_fecun_23 <- ggplot(emmeans_result[trait == "Fecundity" & CGE=="23",], aes(x=Population,y=response, fill = Population))+
  geom_bar(stat = "identity", col="black")+
  scale_fill_manual(values = c("#6e326e","#4fc0c6","#e8a11c"))+
  #geom_errorbar(aes(ymin=10^(lower.CL),ymax=10^(upper.CL)),width=0.2,size=1)+
  geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL), show.legend = FALSE)+
  ylim(0,50)+
  labs(title = "", y="Number of eggs per female", x="Population")+
  geom_signif(comparisons = list(c("Ancestral", "Constant"),
                                 c("Ancestral", "Fluctuating"),
                                 c("Constant", "Fluctuating")),
              annotations = c("***","****","ns"),
              y_position = c(41,47,44))+
  theme(legend.position = "none",
        panel.background = element_rect(fill = alpha("#4fc0c6", 0.3)),
        axis.text = element_text(size = 14),
        axis.title = element_text(size=18),title = element_text(size=20))

p_fecun_2818 <- ggplot(emmeans_result[trait == "Fecundity" & CGE=="2818",], aes(x=Population,y=response, fill = Population))+
  geom_bar(stat = "identity", col="black")+
  scale_fill_manual(values = c("#6e326e","#4fc0c6","#e8a11c"))+
  geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL), show.legend = FALSE)+
  ylim(0,50)+
  labs(title = "", y="Number of eggs per female", x="Population")+
  geom_signif(comparisons = list(c("Ancestral", "Constant"),
                                 c("Ancestral", "Fluctuating"),
                                 c("Constant", "Fluctuating")),
              annotations = c("****","****","ns"),
              y_position = c(41,47,44))+
  theme(legend.position = "none",
        panel.background = element_rect(fill = alpha("#e8a11c", 0.3)),
        axis.text = element_text(size = 14),
        axis.title = element_text(size=18),title = element_text(size=20))

p_relfit_23 <- ggplot(emmeans_result[trait == "Competitive Fitness" & CGE=="23",], aes(x=Population,y=prob, fill = Population))+
  geom_bar(stat = "identity", col="black")+
  scale_fill_manual(values = c("#6e326e","#4fc0c6","#e8a11c"))+
  geom_pointrange(aes(ymin = asymp.LCL, ymax = asymp.UCL), show.legend = FALSE)+
  ylim(0,1)+
  labs(title = "", y=" Proportion of red-eyed flies", x="Population")+
  geom_signif(comparisons = list(c("Ancestral", "Constant"),
                                 c("Ancestral", "Fluctuating"),
                                 c("Constant", "Fluctuating")),
              annotations = c("****","****","ns"),
              y_position = c(0.8,0.95,0.875))+
  theme(legend.position = "none",
        panel.background = element_rect(fill = alpha("#4fc0c6", 0.3)),
        axis.text = element_text(size = 14),
        axis.title = element_text(size=18),title = element_text(size=20))

p_relfit_2818 <- ggplot(emmeans_result[trait == "Competitive Fitness" & CGE=="2818",], aes(x=Population,y=prob, fill = Population))+
  geom_bar(stat = "identity", col="black")+
  scale_fill_manual(values = c("#6e326e","#4fc0c6","#e8a11c"))+
  geom_pointrange(aes(ymin = asymp.LCL, ymax = asymp.UCL), show.legend = FALSE)+
  ylim(0,1)+
  labs(title = "", y="Proportion of red-eyed flies", x="Population")+
  geom_signif(comparisons = list(c("Ancestral", "Constant"),
                                 c("Ancestral", "Fluctuating"),
                                 c("Constant", "Fluctuating")),
              annotations = c("****","****","*"),
              y_position = c(0.8,0.95,0.875))+
  theme(legend.position = "none",
        panel.background = element_rect(fill = alpha("#e8a11c", 0.3)),
        axis.text = element_text(size = 14),
        axis.title = element_text(size=18),title = element_text(size=20))

library(cowplot)
plots <- list(p_fecun_23, p_fecun_2818, p_relfit_23, p_relfit_2818)

merge_p <- plot_grid(
  plotlist = plots, labels = c("A", "B", "C", "D"), label_size = 20,
  nrow=2
)

merge_p_annotated <- ggdraw() +
  draw_plot(merge_p, x = 0.05, y = 0, width = 0.95, height = 0.97) +
  draw_label("23 CGE", x = 0.32, y = 0.95, hjust = 0.5, fontface = "bold", size = 25) +
  draw_label("2818 CGE", x = 0.78, y = 0.95, hjust = 0.5, fontface = "bold", size = 25) +
  draw_label("Fecundity", x = 0.02, y = 0.72, hjust = 0.5, angle = 90, fontface = "bold", size = 25) +
  draw_label("Competitive Fitness", x = 0.02, y = 0.25, hjust = 0.5, angle = 90, fontface = "bold", size = 25) 
  

ggsave("Phenotype_modelfit.pdf", width = 12, height = 10,plot = merge_p_annotated)


```



#4. developmental time
##4.1 23CGE
```{r}
#use the measurement from competitive fitness but check the eclosion trajectory

#read raw data
raw_data<-read_xlsx("competitivefitness/23_competitivefitness_rawdata.xlsx")
pop<-paste(rep(c("Anc_R","Constant_R","wave2_R"), each=5),
           rep(1:5,3),
           sep="")
raw_data<-subset(raw_data, EvoRep %in% pop)

#get accumulative results 
accumu_data<-NULL
for (i in 1:10){
  raw_data1<-subset(raw_data, day %in% 1:i)
  t1<-aggregate(raw_data1$corrected_white, by = list(raw_data1$EvoRep, raw_data1$TechRep), sum) #flies with white eye
  t2<-aggregate(raw_data1$corrected_red, by = list(raw_data1$EvoRep, raw_data1$TechRep), sum) #flies with red eye
  data1<-data.frame(evorep=t1$Group.1, techrep=t1$Group.2, white=t1$x, red=t2$x, Days=i,
                  Population=rep(rep(c("Ancestral","Constant","Fluctuating"),each=5),3))
  accumu_data<-rbind(accumu_data,data1)}
accumu_data$total=accumu_data$red+accumu_data$white
accumu_data$ratio<-accumu_data$red/accumu_data$total
red_day10=subset(accumu_data,Days==10, select = c("evorep","techrep","red"))
colnames(red_day10)=c("evorep","techrep","red_day10")

accumu_data=merge(accumu_data,red_day10,by=c("evorep","techrep"))
accumu_data$ratio_day10=accumu_data$red/accumu_data$red_day10

accumu_plot_data_ratio<-summarySE(accumu_data, measurevar = "ratio_day10", groupvars = c("Population","Days"))
ratio_red<-ggplot(accumu_plot_data_ratio, aes(x=Days, y = ratio_day10, group=Population, col=Population))+
  geom_line()+
  geom_point()+
   annotate("rect", xmin = 0.8, xmax = 1.2, ymin = 0.0, ymax = 0.4, 
           color = "red", fill = NA, linetype = "dashed", size = 0.5) +
  geom_errorbar(aes(ymin=ratio_day10-2*se, ymax=ratio_day10+2*se),width=0.1)+
  xlab("Days")+ylab("Proportion of red-eyed flies")+
  scale_x_continuous(breaks = seq(1,10,1))+
  scale_y_continuous(breaks = seq(0,1,0.2), limits = c(0,1))+
  scale_color_manual(values = c("#6e326e","#4fc0c6","#e8a11c"))+
  ggtitle("23CGE")+
  theme(panel.background = element_rect(fill="#eeeeee"),
        panel.grid.minor = element_blank(),
        legend.position = c(0.1,0.8))

#plot(ratio_red)



accumu_data$Days=factor(accumu_data$Days)
accumu_data$Population=factor(accumu_data$Population)
ratio_red_box<-ggplot(subset(accumu_data,Days==1), aes(x=Population, y = ratio_day10, col=Population))+
  geom_boxplot()+
  ylab("Proportion of red-eyed flies")+
  scale_y_continuous(breaks = seq(0,0.6,0.2), limits = c(0,0.6))+
  scale_color_manual(values = c("#6e326e","#4fc0c6","#e8a11c"))+
  geom_signif(comparisons = list(c("Ancestral","Fluctuating"),c("Ancestral","Constant"),c("Fluctuating", "Constant")),
              test = "wilcox.test",map_signif_level=TRUE,
              step_increase = 0.1,)+
  ggtitle("Day1")+
  theme(panel.background = element_rect(fill="#eeeeee"),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(angle = 0, hjust=0.5))

plot(ratio_red_box)

# Convert the inset plot to a grob
ratio_red_box <- ggplotGrob(ratio_red_box)
merge_plot=ratio_red +
  annotation_custom(
    grob = ratio_red_box,
    xmin = 5, xmax = 10,    # Set position in the x-axis
    ymin = 0.0, ymax = 0.7    # Set position in the y-axis
  )
plot(merge_plot)
ggsave("Competitive_Fitness_trajectory_23CGE.pdf", width = 10, height = 6)
```


##4.2 2818 CGE
```{r}
#use the measurement from competitive fitness but check the eclosion trajectory

#read raw data
raw_data<-read_xlsx("competitivefitness/28_18_competitivefitness_rawdata.xlsx")
pop<-paste(rep(c("Anc_R","Constant_R","wave2_R"), each=5),
           rep(1:5,3),
           sep="")
raw_data<-subset(raw_data, EvoRep %in% pop)

#get accumulative results 
accumu_data<-NULL
for (i in 1:10){
  raw_data1<-subset(raw_data, day %in% 1:i)
  t1<-aggregate(raw_data1$corrected_white, by = list(raw_data1$EvoRep, raw_data1$TechRep), sum) #flies with white eye
  t2<-aggregate(raw_data1$corrected_red, by = list(raw_data1$EvoRep, raw_data1$TechRep), sum) #flies with red eye
  data1<-data.frame(evorep=t1$Group.1, techrep=t1$Group.2, white=t1$x, red=t2$x, Days=i,
                  Population=rep(rep(c("Ancestral","Constant","Fluctuating"),each=5),3))
  accumu_data<-rbind(accumu_data,data1)}
accumu_data$total=accumu_data$red+accumu_data$white
accumu_data$ratio<-accumu_data$red/accumu_data$total
red_day10=subset(accumu_data,Days==10, select = c("evorep","techrep","red"))
colnames(red_day10)=c("evorep","techrep","red_day10")

accumu_data=merge(accumu_data,red_day10,by=c("evorep","techrep"))
accumu_data$ratio_day10=accumu_data$red/accumu_data$red_day10

accumu_plot_data_ratio<-summarySE(accumu_data, measurevar = "ratio_day10", groupvars = c("Population","Days"))
ratio_red<-ggplot(accumu_plot_data_ratio, aes(x=Days, y = ratio_day10, group=Population, col=Population))+
  geom_line()+
  geom_point()+
   annotate("rect", xmin = 0.8, xmax = 1.2, ymin = 0.0, ymax = 0.1, 
           color = "red", fill = NA, linetype = "dashed", size = 0.5) +
  geom_errorbar(aes(ymin=ratio_day10-2*se, ymax=ratio_day10+2*se),width=0.1)+
  xlab("Days")+ylab("Proportion of red-eyed flies")+
  scale_x_continuous(breaks = seq(1,10,1))+
  scale_y_continuous(breaks = seq(0,1,0.2), limits = c(0,1))+
  scale_color_manual(values = c("#6e326e","#4fc0c6","#e8a11c"))+
  ggtitle("2818 CGE")+
  theme(panel.background = element_rect(fill="#eeeeee"),
        panel.grid.minor = element_blank(),
        legend.position = c(0.1,0.8))

#plot(ratio_red)



accumu_data$Days=factor(accumu_data$Days)
accumu_data$Population=factor(accumu_data$Population)
ratio_red_box<-ggplot(subset(accumu_data,Days==1), aes(x=Population, y = ratio_day10, col=Population))+
  geom_boxplot()+
  ylab("Proportion of red-eyed flies")+
  scale_y_continuous(breaks = seq(0,0.12,0.02), limits = c(0,0.12))+
  scale_color_manual(values = c("#6e326e","#4fc0c6","#e8a11c"))+
  geom_signif(comparisons = list(c("Ancestral","Fluctuating"),c("Ancestral","Constant"),c("Fluctuating", "Constant")),
              test = "wilcox.test",map_signif_level=TRUE,
              step_increase = 0.1,)+
  ggtitle("Day1")+
  theme(panel.background = element_rect(fill="#eeeeee"),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(angle = 0, hjust=0.5))

plot(ratio_red_box)

# Convert the inset plot to a grob
ratio_red_box <- ggplotGrob(ratio_red_box)
merge_plot=ratio_red +
  annotation_custom(
    grob = ratio_red_box,
    xmin = 5, xmax = 10,    # Set position in the x-axis
    ymin = 0.0, ymax = 0.7    # Set position in the y-axis
  )
plot(merge_plot)
ggsave("Competitive_Fitness_trajectory_2818CGE.pdf", width = 10, height = 6)
```

