---
title: "compare_match_SNP"
author: "Changyi"
date: "2026-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#this scripts trying to compare the the test target parallele
#e.g. For SNPs that significant in the fluc pop, what are the frequency change of those snp in the cons pop, and are they significantly differ from the match neutral site?
##this scripts update the hunting for the match SNPs, they just the complementary to the test SNPs


```{r}
library(data.table)
library(ggplot2)

```

#functions
```{r}

plot_AFC_distribution <- function(test_afc, match_afc, test_snp=NULL, rep=NULL){

  color_cons<-"#4fc0c6"
  color_fluc<-"#e8a11c"
  color_match<-"grey"
  
  
  #r5
  #focal_afc<-dt_af[loci %in% fluc_only_site$loci,afc_fluc_r5]
 
  
  #check which SNP to test and determined the AFC to plot in which population
  if (test_snp =="Cons"){
     dt_plot<-data.table(
       Significant=c(
         rep("Cons (Test)", length(test_afc)), #AFC of cons sig snps in fluc pop(test)
         rep("Match", length(match_afc))), #AFC of match snps in fluc pop(match)
      AFC = c(test_afc, match_afc)
      )
      
     #dt_plot[, Significant := gsub("Cons", "Cons (Test)", Significant)];
     AFC_pop="Fluc";
     color_test = color_cons;
     back_color=color_fluc}
  else if (test_snp =="Fluc"){
     dt_plot<-data.table(
       Significant=c(
         rep("Fluc (Test)", length(test_afc)), #AFC of fluc sig snps in cons pop(test)
         rep("Match", length(match_afc))), #AFC of match snps in cons pop(match)
       AFC = c(test_afc, match_afc)
       )
     #dt_plot[, Significant := gsub("Fluc", "Fluc (Test)", Significant)]; 
     AFC_pop="Cons";
     color_test = color_fluc;
     back_color=color_cons}
  
  invisible(dt_plot[, Significant := factor(Significant)])
  
  #test if the test data set have greater AFC than the matched SNPs
  p_value<-wilcox.test(match_afc, test_afc, alternative = "less")$p.value
  
  p<- ggplot(dt_plot)+
    geom_density(aes(x=AFC, fill = Significant), alpha=0.5)+ 
    annotate("text", x=0.5, y=Inf, label = paste0("One-sided wilcoxon test\nP-value=",p_value), hjust = 0.6, vjust = 3)+
    scale_fill_manual(values = c(color_test, color_match))+
    labs(title = paste0(AFC_pop, " population Rep",rep))+
    xlim(-1,1)+#ylim(0,25)+
    theme(legend.position = c(0.1,0.8),
          panel.background = element_rect(fill = alpha(back_color,0.2)))
  return(p)
}



```

#1. load data
```{r}

# #load af data
# load("01genomic_data/fluc_HAF.RData")
# dt_af_fluc<-as.data.table(afmat, keep.rownames = TRUE)
# dt_af_fluc <- dt_af_fluc[,grep("rn|anc|fluc\\.60",colnames(dt_af_fluc)), with = FALSE]#keep only F0 and F60
# rm(afmat, samps, sites)
# load("01genomic_data/cons_HAF.RData")
# dt_af_cons<-as.data.table(afmat, keep.rownames = TRUE)
# dt_af_cons <- dt_af_cons[,grep("rn|anc|cons\\.60",colnames(dt_af_cons)), with = FALSE]#keep only F0 and F60
# rm(afmat, samps, sites)
# 
# dt_af<-merge(dt_af_fluc,dt_af_cons, by = c("rn","anc.0.1","anc.0.2","anc.0.3","anc.0.4","anc.0.5"))
# setnames(dt_af, "rn", "loci")
# 
# for (rep in 1:5){
#   dt_af[, paste0("afc_fluc_r",rep) := get(paste0("fluc.60.",rep)) - get(paste0("anc.0.",rep))]
#   dt_af[, paste0("afc_cons_r",rep) := get(paste0("cons.60.",rep)) - get(paste0("anc.0.",rep))]
# }
# dt_af[, mean_fluc_afc := rowMeans(.SD), .SDcols = patterns("^afc_fluc_r")]
# dt_af[, mean_cons_afc := rowMeans(.SD), .SDcols = patterns("^afc_cons_r")]
# fwrite(dt_af, file = "Summary_AFinfo.txt.gz", quote = FALSE, sep="\t")
# rm(dt_af_cons, dt_af_fluc)
```


#2. GLM

##2.1 Test Fluc SNP
```{r}
#snps significant in the fluc population and test in the cons population

#load af information
dt_af<-fread("Summary_AFinfo.txt.gz")

#polarize SNPs to be increase in the fluc population
freq_cols <- grep("^(anc|fluc|cons)", names(dt_af), value = TRUE)
afc_cols <- grep("^(afc|mean)", names(dt_af), value = TRUE)
##rows that need to be flipped
flip_ix <- dt_af$mean_fluc_afc <0
invisible(dt_af[flip_ix, (freq_cols) := lapply(.SD, function(x) 1-x), .SDcols = freq_cols])
invisible(dt_af[flip_ix, (afc_cols) := lapply(.SD, function(x) -x), .SDcols = afc_cols])

##get significant site
fluc_sig_site<-fread("df.sig.fluc.1_7.csv")
fluc_sig_site<-fluc_sig_site[sigLevel>=2,c("chrom","pos")]
invisible(fluc_sig_site[, loci := paste0(chrom,".",pos)])

cons_sig_site<-fread("df.sig.cons.1_7.csv")
cons_sig_site<-cons_sig_site[sigLevel>=2,c("chrom","pos")]
invisible(cons_sig_site[, loci := paste0(chrom,".",pos)])

fluc_only_site<-fluc_sig_site[!(loci %in% cons_sig_site$loci), loci]
cons_only_site<-cons_sig_site[!(loci %in% fluc_sig_site$loci), loci]
sig_site<-union(fluc_sig_site$loci, cons_sig_site$loci)

#check the median allele frequency change across the genome
afc_sig<-median(dt_af[loci %in% fluc_only_site,mean_fluc_afc])
##0.113688

set.seed(1)
summary_afc<-data.table(significant=character(),
                        category=character(),
                        Replicate=numeric(),
                        median_AFC=numeric(),
                        wilcox_test_greater=character(),
                        wilcox_test_less=character(),
                        parallel=character())


for (r in 1:5){
  #extract the replicate
  cols <- c("loci",
            paste0("anc.0.",r),
            paste0("cons.60.",r))  ##here should be the cons population, as we are check fluc sig SNPs in the cons population
  dt_af_rep<-dt_af[, ..cols]
  colnames(dt_af_rep)<-c("loci", "anc", "evo")
  #categorize into 20 bins based on the initial freq
  invisible(dt_af_rep[, freq_bin := cut(anc, 20, labels =FALSE)])

  ##AFC of selection target in it's own environment(constant significant snp in the cons population)
  dt_sig <- dt_af_rep[loci %in% cons_only_site, ]
  sig_afc <- dt_sig$evo - dt_sig$anc
  median_target_afc<-median(sig_afc, na.rm = TRUE)
  
  #test data
  dt_test<-dt_af_rep[loci %in% fluc_only_site,] #test the fluc sig snp in the cons population
  bin_counts <- dt_test[, .N, by = freq_bin]
  test_afc <- dt_test$evo - dt_test$anc
  median_test_afc<-median(test_afc, na.rm = TRUE)
  
  #get matched "neutral" site
  dt_background<- dt_af_rep[!(loci %in% dt_test$loci),] ##remove significant SNP
  match_snp <- bin_counts[, 
                        dt_background[freq_bin == .BY$freq_bin][sample(.N,N)],
                        by = freq_bin] #sample SNPs that match the freq of the test snp
  match_afc<-match_snp$evo - match_snp$anc
  median_match_afc<-median(match_afc, na.rm = TRUE)
  
  #check the consistency
  p_parallel <- wilcox.test(test_afc, match_afc, alternative = "greater")$p.value
  p_anti_parallel <- wilcox.test(test_afc, match_afc, alternative = "less")$p.value
  #p <- t.test(test_afc, match_afc)$p.value
  parallel <- ifelse(p_parallel<=0.05 & p_anti_parallel>0.05, "Parallel",
                     ifelse(p_parallel>0.05 & p_anti_parallel<=0.05, "Anti-Parallel", "No-diff"))

  
  summary_afc <- rbind(summary_afc, list(
    significant=c("fluc", "cons", "cons"),
    category=c("target","test","match"),
    Replicate=c(r,r,r),
    median_AFC=c(median_target_afc,median_test_afc,median_match_afc),
    wilcox_test_greater=c(NA, p_parallel, NA), ##p-value that AFC in the test smaller greater the matched data set
    wilcox_test_less=c(NA, p_anti_parallel, NA), ##p-value that AFC in the test smaller than the matched data set
    parallel=c(NA, parallel,"match")))
  
  
  #plot replicate specific AFC distribution
  p_afc <- plot_AFC_distribution(test_afc, match_afc, test_snp = "Fluc", rep=r)
  ggsave(filename=paste0("plots_compare_match/GLM_AFC_distribution_cons_r",r,".pdf"), plot = p_afc, width = 7, height = 5)
  
}
fwrite(summary_afc, file = "plots_compare_match/GLM_Fluc_sig_SNP_in_Cons_pop.txt", quote = FALSE, sep="\t", na = "NA")

#plot summary results (boxplot)
p_box <- ggplot(summary_afc[category != "target",], aes(x=category, y = median_AFC, color = parallel, shape = category))+
  geom_point(size = 3, position = position_jitter(width = 0.1))+
  ylim(0,0.05)+
  scale_color_manual(
    values = c(
      "Parallel" = "red",
      "Anti-parallel" = "blue",
      "No-diff" = "grey",
      "match" = "black"))+
  scale_shape_manual(values=c(4,16))+
  labs(title="Fluc populations", x="",y="Median AFC")+
  theme_bw()
ggsave(filename = "plots_compare_match/GLM_Fluc_Sig_SNP_in_Cons_pop.pdf", plot = p_box)

####check initial frequency
dt_af_sig <- dt_af[loci %in% fluc_only_site,c("loci","anc.0.1","anc.0.2","anc.0.3","anc.0.4","anc.0.5")]
colnames(dt_af_sig) <- c("loci","rep1","rep2","rep3","rep4","rep5")
af_anc<-melt(dt_af_sig, measure.vars = paste0("rep",1:5), variable.name = "replicate",value.name = "AF")
p_af<-ggplot(af_anc, aes(x=AF, fill=replicate))+
  geom_density(alpha=0.2)+
  facet_wrap(~replicate, nrow=1)+
  theme_minimal()+
  theme(legend.position = "top")
ggsave(filename = "plots_compare_match/GLM_Fluc_sig_SNP_anc_AF_distribution.pdf", plot = p_af)
```



##2.2 Test Cons SNPs

```{r}
##snps significant in the cons population and test in the fluc population

#load af information
dt_af<-fread("Summary_AFinfo.txt.gz")

#polarize SNPs to be increase in the fluc population
freq_cols <- grep("^(anc|fluc|cons)", names(dt_af), value = TRUE)
afc_cols <- grep("^(afc|mean)", names(dt_af), value = TRUE)
##rows that need to be flipped
flip_ix <- dt_af$mean_cons_afc <0
invisible(dt_af[flip_ix, (freq_cols) := lapply(.SD, function(x) 1-x), .SDcols = freq_cols])
invisible(dt_af[flip_ix, (afc_cols) := lapply(.SD, function(x) -x), .SDcols = afc_cols])

##get significant site
fluc_sig_site<-fread("df.sig.fluc.1_7.csv")
fluc_sig_site<-fluc_sig_site[sigLevel>=2,c("chrom","pos")]
invisible(fluc_sig_site[, loci := paste0(chrom,".",pos)])

cons_sig_site<-fread("df.sig.cons.1_7.csv")
cons_sig_site<-cons_sig_site[sigLevel>=2,c("chrom","pos")]
invisible(cons_sig_site[, loci := paste0(chrom,".",pos)])

fluc_only_site<-fluc_sig_site[!(loci %in% cons_sig_site$loci), loci]
cons_only_site<-cons_sig_site[!(loci %in% fluc_sig_site$loci), loci]
sig_site<-union(fluc_sig_site$loci, cons_sig_site$loci)

#check the median allele frequency change across the genome
afc_sig<-median(dt_af[loci %in% cons_only_site,mean_cons_afc])
##0.123868

set.seed(1)
summary_afc<-data.table(significant=character(),
                        category=character(),
                        Replicate=numeric(),
                        median_AFC=numeric(),
                        wilcox_test_greater=character(),
                        wilcox_test_less=character(),
                        parallel=character())


for (r in 1:5){
  #extract the replicate
  cols <- c("loci",
            paste0("anc.0.",r),
            paste0("fluc.60.",r))  ##here should be the fluc population, as we are check cons sig SNPs in the fluc population
  dt_af_rep<-dt_af[, ..cols]
  colnames(dt_af_rep)<-c("loci", "anc", "evo")
  #categorize into 20 bins based on the initial freq
  invisible(dt_af_rep[, freq_bin := cut(anc, 20, labels =FALSE)])

  ##AFC of selection target in it's own environment(fluctuating significant snp in the fluc population)
  dt_sig <- dt_af_rep[loci %in% fluc_only_site, ]
  sig_afc <- dt_sig$evo - dt_sig$anc
  median_target_afc<-median(sig_afc, na.rm = TRUE)
  
  #test data
  dt_test<-dt_af_rep[loci %in% cons_only_site,]
  bin_counts <- dt_test[, .N, by = freq_bin]
  test_afc <- dt_test$evo - dt_test$anc
  median_test_afc<-median(test_afc, na.rm = TRUE)
  
  #get matched "neutral" site
  dt_background<- dt_af_rep[!(loci %in% dt_test$loci),] ##remove significant SNP
  match_snp <- bin_counts[, 
                        dt_background[freq_bin == .BY$freq_bin][sample(.N,N)],
                        by = freq_bin] #sample SNPs that match the freq
  match_afc<-match_snp$evo - match_snp$anc
  median_match_afc<-median(match_afc, na.rm = TRUE)
  
  #check the consistency
  p_parallel <- wilcox.test(test_afc, match_afc, alternative = "greater")$p.value
  p_anti_parallel <- wilcox.test(test_afc, match_afc, alternative = "less")$p.value
  #p <- t.test(test_afc, match_afc)$p.value
  parallel <- ifelse(p_parallel<=0.05 & p_anti_parallel>0.05, "Parallel",
                     ifelse(p_parallel>0.05 & p_anti_parallel<=0.05, "Anti-Parallel", "No-diff"))

  
  summary_afc <- rbind(summary_afc, list(
    significant=c("fluc", "cons", "cons"),
    category=c("target","test","match"),
    Replicate=c(r,r,r),
    median_AFC=c(median_target_afc,median_test_afc,median_match_afc),
    wilcox_test_greater=c(NA, p_parallel, NA), ##p-value that AFC in the test smaller greater the matched data set
    wilcox_test_less=c(NA, p_anti_parallel, NA), ##p-value that AFC in the test smaller than the matched data set
    parallel=c(NA, parallel,"match")))
  
  
  #plot replicate specific AFC distribution
  p_afc <- plot_AFC_distribution(test_afc, match_afc, test_snp = "Cons", rep=r)
  ggsave(filename=paste0("plots_compare_match/GLM_AFC_distribution_fluc_r",r,".pdf"), plot = p_afc, width = 7, height = 5)
  
}
fwrite(summary_afc, file = "plots_compare_match/GLM_Cons_sig_SNP_in_Fluc_pop.txt", quote = FALSE, sep="\t", na = "NA")

#plot summary results (boxplot)
p_box <- ggplot(summary_afc[category != "target",], aes(x=category, y = median_AFC, color = parallel, shape = category))+
  geom_point(size = 3, position = position_jitter(width = 0.1))+
  ylim(0,0.05)+
  scale_color_manual(
    values = c(
      "Parallel" = "red",
      "Anti-parallel" = "blue",
      "No-diff" = "grey",
      "match" = "black"))+
  scale_shape_manual(values=c(4,16))+
  labs(title="Fluc populations", x="",y="Median AFC")+
  theme_bw()
ggsave(filename = "plots_compare_match/GLM_Cons_Sig_SNP_in_Fluc_pop.pdf", plot = p_box)

####check initial frequency
dt_af_sig <- dt_af[loci %in% fluc_only_site,c("loci","anc.0.1","anc.0.2","anc.0.3","anc.0.4","anc.0.5")]
colnames(dt_af_sig) <- c("loci","rep1","rep2","rep3","rep4","rep5")
af_anc<-melt(dt_af_sig, measure.vars = paste0("rep",1:5), variable.name = "replicate",value.name = "AF")
p_af<-ggplot(af_anc, aes(x=AF, fill=replicate))+
  geom_density(alpha=0.2)+
  facet_wrap(~replicate, nrow=1)+
  theme_minimal()+
  theme(legend.position = "top")
ggsave(filename = "plots_compare_match/GLM_Cons_sig_SNP_anc_AF_distribution.pdf", plot = p_af)
```

#3. adapted CMH

##3.1 Test Fluc SNP
```{r}
#snps significant in the fluc population and test in the cons population

#load af information
dt_af<-fread("Summary_AFinfo.txt.gz")

#polarize SNPs to be increase in the fluc population
freq_cols <- grep("^(anc|fluc|cons)", names(dt_af), value = TRUE)
afc_cols <- grep("^(afc|mean)", names(dt_af), value = TRUE)
##rows that need to be flipped
flip_ix <- dt_af$mean_fluc_afc <0
invisible(dt_af[flip_ix, (freq_cols) := lapply(.SD, function(x) 1-x), .SDcols = freq_cols])
invisible(dt_af[flip_ix, (afc_cols) := lapply(.SD, function(x) -x), .SDcols = afc_cols])

##get significant site
fluc_sig_site<-fread("CMH_fluc.result.gz")
fluc_sig_site<-fluc_sig_site[adj_p<=0.05,c("chr","pos")]
invisible(fluc_sig_site[, loci := paste0(chr,".",pos)])

cons_sig_site<-fread("CMH_cons.result.gz")
cons_sig_site<-cons_sig_site[adj_p<=0.05,c("chr","pos")]
invisible(cons_sig_site[, loci := paste0(chr,".",pos)])

fluc_only_site<-fluc_sig_site[!(loci %in% cons_sig_site$loci), loci]
cons_only_site<-cons_sig_site[!(loci %in% fluc_sig_site$loci), loci]
sig_site<-union(fluc_sig_site$loci, cons_sig_site$loci)

#check the median allele frequency change across the genome
afc_sig<-median(dt_af[loci %in% fluc_only_site,mean_fluc_afc], na.rm = TRUE)
##0.16209

set.seed(1)
summary_afc<-data.table(significant=character(),
                        category=character(),
                        Replicate=numeric(),
                        median_AFC=numeric(),
                        wilcox_test_greater=character(),
                        wilcox_test_less=character(),
                        parallel=character())


for (r in 1:5){
  #extract the replicate
  cols <- c("loci",
            paste0("anc.0.",r),
            paste0("cons.60.",r))  ##here should be the cons population, as we are check fluc sig SNPs in the cons population
  dt_af_rep<-dt_af[, ..cols]
  colnames(dt_af_rep)<-c("loci", "anc", "evo")
  #categorize into 20 bins based on the initial freq
  invisible(dt_af_rep[, freq_bin := cut(anc, 20, labels =FALSE)])

  ##AFC of selection target in it's own environment(constant significant snp in the cons population)
  dt_sig <- dt_af_rep[loci %in% cons_only_site, ]
  sig_afc <- dt_sig$evo - dt_sig$anc
  median_target_afc<-median(sig_afc, na.rm = TRUE)
  
  #test data
  dt_test<-dt_af_rep[loci %in% fluc_only_site,] #test the fluc sig snp in the cons population
  bin_counts <- dt_test[, .N, by = freq_bin]
  test_afc <- dt_test$evo - dt_test$anc
  median_test_afc<-median(test_afc, na.rm = TRUE)
  
  #get matched "neutral" site
  dt_background<- dt_af_rep[!(loci %in% dt_test$loci),] ##remove significant SNP
  match_snp <- bin_counts[, 
                        dt_background[freq_bin == .BY$freq_bin][sample(.N,N)],
                        by = freq_bin] #sample SNPs that match the freq of the test snp
  match_afc<-match_snp$evo - match_snp$anc
  median_match_afc<-median(match_afc, na.rm = TRUE)
  
  #check the consistency
  p_parallel <- wilcox.test(test_afc, match_afc, alternative = "greater")$p.value
  p_anti_parallel <- wilcox.test(test_afc, match_afc, alternative = "less")$p.value
  #p <- t.test(test_afc, match_afc)$p.value
  parallel <- ifelse(p_parallel<=0.05 & p_anti_parallel>0.05, "Parallel",
                     ifelse(p_parallel>0.05 & p_anti_parallel<=0.05, "Anti-Parallel", "No-diff"))

  
  summary_afc <- rbind(summary_afc, list(
    significant=c("fluc", "cons", "cons"),
    category=c("target","test","match"),
    Replicate=c(r,r,r),
    median_AFC=c(median_target_afc,median_test_afc,median_match_afc),
    wilcox_test_greater=c(NA, p_parallel, NA), ##p-value that AFC in the test smaller greater the matched data set
    wilcox_test_less=c(NA, p_anti_parallel, NA), ##p-value that AFC in the test smaller than the matched data set
    parallel=c(NA, parallel,"match")))
  
  
  #plot replicate specific AFC distribution
  p_afc <- plot_AFC_distribution(test_afc, match_afc, test_snp = "Fluc", rep=r)
  ggsave(filename=paste0("plots_compare_match/CMH_AFC_distribution_cons_r",r,".pdf"), plot = p_afc, width = 7, height = 5)
  
}
fwrite(summary_afc, file = "plots_compare_match/CMH_Fluc_sig_SNP_in_Cons_pop.txt", quote = FALSE, sep="\t", na = "NA")

#plot summary results (boxplot)
p_box <- ggplot(summary_afc[category != "target",], aes(x=category, y = median_AFC, color = parallel, shape = category))+
  geom_point(size = 3, position = position_jitter(width = 0.1))+
  #ylim(0,0.05)+
  scale_color_manual(
    values = c(
      "Parallel" = "red",
      "Anti-parallel" = "blue",
      "No-diff" = "grey",
      "match" = "black"))+
  scale_shape_manual(values=c(4,16))+
  labs(title="Cons populations", x="",y="Median AFC")+
  theme_bw()
ggsave(filename = "plots_compare_match/CMH_Fluc_Sig_SNP_in_Cons_pop.pdf", plot = p_box)

####check initial frequency
dt_af_sig <- dt_af[loci %in% fluc_only_site,c("loci","anc.0.1","anc.0.2","anc.0.3","anc.0.4","anc.0.5")]
colnames(dt_af_sig) <- c("loci","rep1","rep2","rep3","rep4","rep5")
af_anc<-melt(dt_af_sig, measure.vars = paste0("rep",1:5), variable.name = "replicate",value.name = "AF")
p_af<-ggplot(af_anc, aes(x=AF, fill=replicate))+
  geom_density(alpha=0.2)+
  facet_wrap(~replicate, nrow=1)+
  theme_minimal()+
  theme(legend.position = "top")
ggsave(filename = "plots_compare_match/CMH_Fluc_sig_SNP_anc_AF_distribution.pdf", plot = p_af)
```


##3.2 Test Cons SNPs

```{r}
##snps significant in the cons population and test in the fluc population

#load af information
dt_af<-fread("Summary_AFinfo.txt.gz")

#polarize SNPs to be increase in the fluc population
freq_cols <- grep("^(anc|fluc|cons)", names(dt_af), value = TRUE)
afc_cols <- grep("^(afc|mean)", names(dt_af), value = TRUE)
##rows that need to be flipped
flip_ix <- dt_af$mean_cons_afc <0
invisible(dt_af[flip_ix, (freq_cols) := lapply(.SD, function(x) 1-x), .SDcols = freq_cols])
invisible(dt_af[flip_ix, (afc_cols) := lapply(.SD, function(x) -x), .SDcols = afc_cols])

##get significant site
fluc_sig_site<-fread("CMH_fluc.result.gz")
fluc_sig_site<-fluc_sig_site[adj_p<=0.05,c("chr","pos")]
invisible(fluc_sig_site[, loci := paste0(chr,".",pos)])

cons_sig_site<-fread("CMH_cons.result.gz")
cons_sig_site<-cons_sig_site[adj_p<=0.05,c("chr","pos")]
invisible(cons_sig_site[, loci := paste0(chr,".",pos)])

fluc_only_site<-fluc_sig_site[!(loci %in% cons_sig_site$loci), loci]
cons_only_site<-cons_sig_site[!(loci %in% fluc_sig_site$loci), loci]
sig_site<-union(fluc_sig_site$loci, cons_sig_site$loci)


#check the median allele frequency change across the genome
afc_sig<-median(dt_af[loci %in% cons_only_site,mean_cons_afc])
##0.1200159

set.seed(1)
summary_afc<-data.table(significant=character(),
                        category=character(),
                        Replicate=numeric(),
                        median_AFC=numeric(),
                        wilcox_test_greater=character(),
                        wilcox_test_less=character(),
                        parallel=character())


for (r in 1:5){
  #extract the replicate
  cols <- c("loci",
            paste0("anc.0.",r),
            paste0("fluc.60.",r))  ##here should be the fluc population, as we are check cons sig SNPs in the fluc population
  dt_af_rep<-dt_af[, ..cols]
  colnames(dt_af_rep)<-c("loci", "anc", "evo")
  #categorize into 20 bins based on the initial freq
  invisible(dt_af_rep[, freq_bin := cut(anc, 20, labels =FALSE)])

  ##AFC of selection target in it's own environment(fluc significant snp in the fluc population)
  dt_sig <- dt_af_rep[loci %in% fluc_only_site, ]
  sig_afc <- dt_sig$evo - dt_sig$anc
  median_target_afc<-median(sig_afc, na.rm = TRUE)
  
  #test data
  dt_test<-dt_af_rep[loci %in% cons_only_site,]
  bin_counts <- dt_test[, .N, by = freq_bin]
  test_afc <- dt_test$evo - dt_test$anc
  median_test_afc<-median(test_afc, na.rm = TRUE)
  
  #get matched "neutral" site
  dt_background<- dt_af_rep[!(loci %in% dt_test$loci),] ##remove significant SNP
  match_snp <- bin_counts[, 
                        dt_background[freq_bin == .BY$freq_bin][sample(.N,N)],
                        by = freq_bin] #sample SNPs that match the freq
  match_afc<-match_snp$evo - match_snp$anc
  median_match_afc<-median(match_afc, na.rm = TRUE)
  
  #check the consistency
  p_parallel <- wilcox.test(test_afc, match_afc, alternative = "greater")$p.value
  p_anti_parallel <- wilcox.test(test_afc, match_afc, alternative = "less")$p.value
  #p <- t.test(test_afc, match_afc)$p.value
  parallel <- ifelse(p_parallel<=0.05 & p_anti_parallel>0.05, "Parallel",
                     ifelse(p_parallel>0.05 & p_anti_parallel<=0.05, "Anti-Parallel", "No-diff"))

  
  summary_afc <- rbind(summary_afc, list(
    significant=c("fluc", "cons", "cons"),
    category=c("target","test","match"),
    Replicate=c(r,r,r),
    median_AFC=c(median_target_afc,median_test_afc,median_match_afc),
    wilcox_test_greater=c(NA, p_parallel, NA), ##p-value that AFC in the test smaller greater the matched data set
    wilcox_test_less=c(NA, p_anti_parallel, NA), ##p-value that AFC in the test smaller than the matched data set
    parallel=c(NA, parallel,"match")))
  
  
  #plot replicate specific AFC distribution
  p_afc <- plot_AFC_distribution(test_afc, match_afc, test_snp = "Cons", rep=r)
  ggsave(filename=paste0("plots_compare_match/CMH_AFC_distribution_fluc_r",r,".pdf"), plot = p_afc, width = 7, height = 5)
  
}
fwrite(summary_afc, file = "plots_compare_match/CMH_Cons_sig_SNP_in_Fluc_pop.txt", quote = FALSE, sep="\t", na = "NA")

#plot summary results (boxplot)
p_box <- ggplot(summary_afc[category != "target",], aes(x=category, y = median_AFC, color = parallel, shape = category))+
  geom_point(size = 3, position = position_jitter(width = 0.1))+
  #ylim(0,0.05)+
  scale_color_manual(
    values = c(
      "Parallel" = "red",
      "Anti-parallel" = "blue",
      "No-diff" = "grey",
      "match" = "black"))+
  scale_shape_manual(values=c(4,16))+
  labs(title="Fluc populations", x="",y="Median AFC")+
  theme_bw()
ggsave(filename = "plots_compare_match/CMH_Cons_Sig_SNP_in_Fluc_pop.pdf", plot = p_box)

####check initial frequency
dt_af_sig <- dt_af[loci %in% fluc_only_site,c("loci","anc.0.1","anc.0.2","anc.0.3","anc.0.4","anc.0.5")]
colnames(dt_af_sig) <- c("loci","rep1","rep2","rep3","rep4","rep5")
af_anc<-melt(dt_af_sig, measure.vars = paste0("rep",1:5), variable.name = "replicate",value.name = "AF")
p_af<-ggplot(af_anc, aes(x=AF, fill=replicate))+
  geom_density(alpha=0.2)+
  facet_wrap(~replicate, nrow=1)+
  theme_minimal()+
  theme(legend.position = "top")
ggsave(filename = "plots_compare_match/CMH_cons_sig_SNP_anc_AF_distribution.pdf", plot = p_af)
```